---
title: "Exploratory Analysis"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
cleaned_wfls = 
  read_csv("./data/cleaned_wfls.csv")

zipcodes =
  read_xlsx("./data/zip_codes.xlsx") %>% 
  mutate(zipcode = as.character(zipcode))

merged_wfls = 
  left_join(cleaned_wfls, zipcodes, by = "zipcode")

write.csv(merged_wfls, "./data/merged_wfls.csv")

merged_wfls
```

### Leave and Race


```{r}

  leave =
  merged_wfls %>% 
  drop_na(race) %>% 
  drop_na(leave_weeks) %>% 
 mutate(race = fct_reorder(race, leave_weeks)) %>% 
  ggplot(aes(x = race, y = leave_weeks, fill = race, alpha = 0.9)) + 
    geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1), legend.position = "none") +
  labs(
    title = "Weeks of maternity leave by race",
    x = "Race",
    y = "Weeks of leave"
  )

ggplotly(leave)
```

```{r}
unpaid = 
  merged_wfls %>% 
  drop_na(race) %>% 
  drop_na(unpaid_leave_weeks) %>% 
  drop_na(leave_weeks) %>% 
 mutate(race = fct_reorder(race, leave_weeks)) %>% 
  ggplot(aes(x = race, y = unpaid_leave_weeks, fill = race, alpha = 0.9)) + 
    geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1), legend.position = "none") +
  ylim(0, 50) + 
  labs(
    title = "Unpaid maternity leave by race",
    x = "Race",
    y = "Unpaid Weeks of leave"
  )

ggplotly(unpaid)
```


The visualizations illustrate that weeks of leave do not substantially differ by race. For paid and unpaid leave, mothers who identify as American Indian/AN take slightly more weeks of leave. Mothers who identify as American Indian/AN also have higher spread of length of leave. 


### Leave and Parenting Status
```{r}
parent = 
  merged_wfls %>% 
  drop_na(partner) %>% 
  drop_na(leave_weeks) %>%
 ggplot(aes(x = partner, y = leave_weeks, fill = partner, alpha = 0.9)) +
  geom_boxplot() +
  labs(
    title = "Weeks of maternity leave by co-parenting status",
    x = "Co-parenting status",
    y = "Weeks of leave"
  )
ggplotly(parent)
```

This visualization shows that there is not a substantial difference of length of leave between mothers who are co-parents and single parents. 


### Leave and Income

```{r}
unpaid = 
merged_wfls %>% 
drop_na(family_income) %>% 
drop_na(unpaid_leave_weeks) %>% 
mutate(food_insecurity = as.character(food_insecurity), 
       food_insecurity = case_when(
         food_insecurity == '0' ~ "no", 
         food_insecurity == '1' ~ "yes"), 
       family_income = fct_reorder(family_income, es3)) %>% 
   ggplot(aes(x = family_income, y = unpaid_leave_weeks, color = food_insecurity)) +
  geom_point() + 
theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) +
  ylim(0, 80) + 
  labs(
    title = "Unpaid maternity leave by income",
    x = "Family income",
    y = "Weeks of unpaid leave", 
    color = "Food insecurity"
  )

ggplotly(unpaid)

```

This plot shows that mothers who are food insecure are clustered towards the lower side of family income. Mothers who are food insecure tend to have a family income less than 40,000. Mothers whose family income is between 16,001 and 20,000 tend to have the shortest amount of unpaid leave. 




### Unpaid Leave and Job Type
```{r}
job_type = 
  merged_wfls %>%  
  drop_na(job_type) %>% 
  drop_na(unpaid_leave_weeks) %>% 
  ggplot(aes(x = job_type, y = unpaid_leave_weeks, fill = job_type, alpha = 0.9)) + 
  geom_boxplot() + 
  ylim(0, 50) + 
  labs(
    title = "Unpaid Leave Weeks by Job Type",
    x = "Job type",
    y = "Unpaid Leave Weeks") + 
  theme(legend.position = "none")

ggplotly(job_type)
```

This visualization shows that mothers who were self-employed were slightly more likely to have longer unpaid leave weeks. However, these values did not value substantially across groups. 

### Returning to Work

```{r}
unemploy = 
merged_wfls %>% 
  drop_na(unemploy_reason) %>% 
  ggplot(aes(y = unemploy_reason, fill = race)) + geom_bar(position = "dodge") + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Reasons for Unemployment",
    x = "Count",
    y = "Reason") 

ggplotly(unemploy)
```


```{r}
unemploy2 = 
merged_wfls %>% 
  drop_na(unemploy_reason) %>% 
  drop_na(education) %>% 
   mutate(education = fct_reorder(education, d7)) %>% 
  ggplot(aes(y = unemploy_reason, fill = education)) + geom_bar(position = "dodge") + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Reasons for Unemployment",
    x = "Count",
    y = "Reason") 

ggplotly(unemploy2)
```


These visualizations shed light on the reason mothers were not back at work at the time of the survey. Choosing to stay at home was the most common response. However, it was disproportionately cited by mothers who were white and mothers who had a 4 year education or higher. 

### Post Partum Check Up Status

```{r}
check = 
merged_wfls %>% 
  drop_na(family_income) %>% 
  drop_na(postpartum_check) %>% 
  mutate( 
     family_income = fct_reorder(family_income, es3)) %>% 
  ggplot(aes(x = family_income, fill = postpartum_check)) + geom_bar(position = "stack") + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Postpartum Checkup Status by Income",
    x = "Family Income",
    y = "Count", 
    fill = "Checkup Status") 

ggplotly(check)
```

Across all income categories, over 90% of mothers received a postpartum check up 4-6 weeks after giving birth. Mothers with a lower family income were more likely to not get a postpartum check up.


These visualizations suggest there were not stark differences of leave by sociodemographic characteristics. However, New York City is a very heterogeneous city, with many pockets of inequality in close proximity. We then wanted to look at leave visually across zipcode. You can find that analysis here.  