---
title: "Project Report Draft"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(tidyverse)
library(readxl)
library(plotly)
library(janitor)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%", 
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

<style type="text/css">

h1.title {
  text-align: center;
}

</style>


## Questions

1. How does length of maternity leave taken by women in New York City vary according to key socioeconomic characteristics?

2. How does length of maternity leave differ by zip code? Do these spacial patterns match the distribution of key socioeconomic characteristics?

3. Is the geographic distribution of average length of maternity leave associated with X, Y, X?

* note: fill in 

## Data {.tabset}

Our data was taken from the New York City Work and Family Leave Survey [WFLS](https://data.cityofnewyork.us/Health/New-York-City-Work-and-Family-Leave-Survey-WFLS-20/grnn-mvqe). It was conducted in March 2016 and was a telephone survey of New York City residents who gave birth in 2014. Its goal was to improve understanding about the availability and accessibility of paid family leave to working parents. The WFLS also sought to describe the role that paid family leave policies play in achieving health equity for parents and children. Additionally, it played a role in the passage of New York's Paid Family Leave Law. A link to the Paid Family Leave Report can be found [here](https://www1.nyc.gov/assets/doh/downloads/pdf/hca/paid-family-leave-report1.pdf). 

We obtained the data from the [NYC Open Data](https://data.cityofnewyork.us/Health/New-York-City-Work-and-Family-Leave-Survey-WFLS-20/grnn-mvqe). From here, we exported the data as a csv file. The raw dataset contained 1063 rows, and 153 columns. We created new variables based on the questionnaire about variables of interest. Our new variable names described the question and the responses described the responses, rather than the numeric responses in the survey questionnaire. his data was then merged with zipcode shape files of New York City.

Our cleaned data consists of 1,063 rows and 20 columns. It includes information on the variables below. 

*Note: How did we make our outcome? 
* NOTE: add something about shape data files. where obtained, how was it cleaned? 


### **Variables of interest***

**Outcomes**

 * `leave_weeks` = weeks of maternity leave (paid or unpaid)
 * `paid_leave_weeks` = weeks of paid maternity leave 
 * `unpaid_leave_weeks` = weeks of unpaid maternity leave
 * `postpartum_check` = had a postpartum checkup 4-6 weeks after giving birth
 
**Predictors**

 * `neighborhood` = neighborhood (based on five digit zip code)
 * `race` = race of survey respondent
 * `family_income` = household family income
 * `food_insecurity` = concern about having enough food to feed family or receives public assistance in the form of SNAP benefits / food stamps
 * `education` = highest level of school completed
 * `partner` = co-parenting status (single or co-parent)
 * `employment_type` = type of employment during pregnancy (government, private company, non-profit, self-employed)


*NOTE: does this need to be here? 

```{r, include=FALSE}
cleaned_wfls = 
  read_csv("./data/cleaned_wfls.csv")

zipcodes =
  read_xlsx("./data/zip_codes.xlsx") %>% 
  mutate(zipcode = as.character(zipcode))

merged_wfls = 
  left_join(cleaned_wfls, zipcodes, by = "zipcode")

write.csv(merged_wfls, "./data/merged_wfls.csv")

merged_wfls
```




As this will be your only chance to describe your project in detail, make sure that your report is a standalone document that fully describes your process and results. 
justify steps you took. major changes to ideas 



## Exploratory Analysis 


### Descriptive Analysis 

We first wanted to examine the distribution of some of our key variables. This helped us better understand the sample. 


#### Length of Leave


```{r}
leave_weeks = 
merged_wfls %>% 
  drop_na(leave_weeks) %>% 
  ggplot(aes(x = leave_weeks)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Fig 1. Leave weeks",
    x = "Number of Weeks",
    y = "Count") 

leave_weeks
```


```{r}
unpaid_leave_weeks = 
merged_wfls %>% 
  drop_na(unpaid_leave_weeks) %>% 
  ggplot(aes(x = unpaid_leave_weeks)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Fig 2. Unpaid leave weeks",
    x = "Number of Weeks",
    y = "Count") 

unpaid_leave_weeks

```

Figure 1 shows the spread of total leave weeks. The most common value was 12 weeks. The spread of time of leave is more likely to be under 12 weeks. Figure 2  illustrates that over 200 mothers did not take unpaid leave time. We were really interested understanding the patterns of the 447 mothers who reported taking unpaid leave. 


#### Race, Education, and Income
```{r}
race = 
merged_wfls %>% 
  drop_na(race) %>% 
  ggplot(aes(x = race)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Fig 3. Distribution of Race",
    x = "Self identified race",
    y = "Count") 

race
```

```{r}
education = 
merged_wfls %>% 
  drop_na(education) %>% 
   mutate(education = fct_reorder(education, d7)) %>% 
  ggplot(aes(x = education)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Fig 4. Distribution of education",
    x = "education level",
    y = "Count") 

education
```


```{r}
income = 
merged_wfls %>% 
  drop_na(family_income) %>% 
  ggplot(aes(x = family_income)) + geom_bar() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Fig 5. Distribution of income",
    x = "income level",
    y = "Count") 

income
```


Figure 3 illustrates that white mothers were over represented in this sample. Figure 4 illustrates that mothers with a four year college degree or higher had higher representation in this data. Lastly, Figure 5 shows that those in households of 85,000 dollars or more were most sampled. However, a limitation of this categorization is that there may be a lot of heterogeneity in the highest income category, especially in an expensive city like New York. 


After getting a sense of the distribution across key variables, our exploratory analysis aimed to examine how the length of leave varied by sociodemographic information. We looked at paid and total length of leave by race, leave by parenting status, leave by job type, and leave by income. We also looked at reasons for unemployment across race and education. Lastly, we examined postpartum checkup status by income. You can find that analysis here. 

* NOTE: about maps 



## Spatial Regression and Analysis 
If you undertake formal statistical analyses, describe these in detail

## Discussion 

* What were findings
      * did not find any differences of leave (paid or unpaid) by race, parenting status, or income. This was not expected. 
      * Reasons for unemployment, interesting that white mothers and mothers with a for year college degree or higher were most likely to choose to stay at home. highlights inequalities in society. 
* what you expected? 
* what insights into the data can you make?
      * our data was almost exclusively categorical. 
      * look at more representative sample of NYC
      * look at paternal leave 


Note: Take out conclusion tab 




**NOTE: on new tab**

## Exploratory Anlaysis  

### Leave and Race


```{r}

  leave =
  merged_wfls %>% 
  drop_na(race) %>% 
  drop_na(leave_weeks) %>% 
 mutate(race = fct_reorder(race, leave_weeks)) %>% 
  ggplot(aes(x = race, y = leave_weeks, fill = race, alpha = 0.9)) + 
    geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1), legend.position = "none") +
  labs(
    title = "Weeks of maternity leave by race",
    x = "Race",
    y = "Weeks of leave"
  )

ggplotly(leave)
```

```{r}
unpaid = 
  merged_wfls %>% 
  drop_na(race) %>% 
  drop_na(unpaid_leave_weeks) %>% 
  drop_na(leave_weeks) %>% 
 mutate(race = fct_reorder(race, leave_weeks)) %>% 
  ggplot(aes(x = race, y = unpaid_leave_weeks, fill = race, alpha = 0.9)) + 
    geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1), legend.position = "none") +
  ylim(0, 50) + 
  labs(
    title = "Unpaid maternity leave by race",
    x = "Race",
    y = "Unpaid Weeks of leave"
  )

ggplotly(unpaid)
```


The visualizations illustrate that weeks of leave do not substantially differ by race. For paid and unpaid leave, mothers who identify as American Indian/AN take slightly more weeks of leave. Mothers who identify as American Indian/AN also have higher spread of length of leave. 


### Leave and Parenting Status
```{r}
parent = 
  merged_wfls %>% 
  drop_na(partner) %>% 
  drop_na(leave_weeks) %>%
 ggplot(aes(x = partner, y = leave_weeks, fill = partner, alpha = 0.9)) +
  geom_boxplot() +
  labs(
    title = "Weeks of maternity leave by co-parenting status",
    x = "Co-parenting status",
    y = "Weeks of leave"
  )
ggplotly(parent)
```

This visualization shows that there is not a substantial difference of length of leave between mothers who are co-parents and single parents. 


### Leave and Income

```{r}
unpaid = 
merged_wfls %>% 
drop_na(family_income) %>% 
drop_na(unpaid_leave_weeks) %>% 
mutate(food_insecurity = as.character(food_insecurity), 
       food_insecurity = case_when(
         food_insecurity == '0' ~ "no", 
         food_insecurity == '1' ~ "yes"), 
       family_income = fct_reorder(family_income, es3)) %>% 
   ggplot(aes(x = family_income, y = unpaid_leave_weeks, color = food_insecurity)) +
  geom_point() + 
theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) +
  ylim(0, 80) + 
  labs(
    title = "Unpaid maternity leave by income",
    x = "Family income",
    y = "Weeks of unpaid leave", 
    color = "Food insecurity"
  )

ggplotly(unpaid)

```

This plot shows that mothers who are food insecure are clustered towards the lower side of family income. Mothers who are food insecure tend to have a family income less than 40,000. Mothers whose family income is between 16,001 and 20,000 tend to have the shortest amount of unpaid leave. 




### Unpaid Leave and Job Type
```{r}
job_type = 
  merged_wfls %>%  
  drop_na(job_type) %>% 
  drop_na(unpaid_leave_weeks) %>% 
  ggplot(aes(x = job_type, y = unpaid_leave_weeks, fill = job_type, alpha = 0.9)) + 
  geom_boxplot() + 
  ylim(0, 50) + 
  labs(
    title = "Unpaid Leave Weeks by Job Type",
    x = "Job type",
    y = "Unpaid Leave Weeks") + 
  theme(legend.position = "none")

ggplotly(job_type)
```

This visualization shows that mothers who were self-employed were slightly more likely to have longer unpaid leave weeks. However, these values did not value substantially across groups. 

### Returning to Work

```{r}
unemploy = 
merged_wfls %>% 
  drop_na(unemploy_reason) %>% 
  ggplot(aes(y = unemploy_reason, fill = race)) + geom_bar(position = "dodge") + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Reasons for Unemployment",
    x = "Count",
    y = "Reason") 

ggplotly(unemploy)
```


```{r}
unemploy2 = 
merged_wfls %>% 
  drop_na(unemploy_reason) %>% 
  drop_na(education) %>% 
   mutate(education = fct_reorder(education, d7)) %>% 
  ggplot(aes(y = unemploy_reason, fill = education)) + geom_bar(position = "dodge") + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Reasons for Unemployment",
    x = "Count",
    y = "Reason") 

ggplotly(unemploy2)
```


These visualizations shed light on the reason mothers were not back at work at the time of the survey. Choosing to stay at home was the most common response. However, it was disproportionately cited by mothers who were white and mothers who had a 4 year education or higher. 

### Post Partum Check Up Status

```{r}
check = 
merged_wfls %>% 
  drop_na(family_income) %>% 
  drop_na(postpartum_check) %>% 
  mutate( 
     family_income = fct_reorder(family_income, es3)) %>% 
  ggplot(aes(x = family_income, fill = postpartum_check)) + geom_bar(position = "stack") + 
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1)) + 
  labs(
    title = "Postpartum Checkup Status by Income",
    x = "Family Income",
    y = "Count", 
    fill = "Checkup Status") 

ggplotly(check)
```

note: made proportion but looked like last hw and wasn't exciting b/c all well over .9

Across all income categories, over 90% of mothers received a postpartum check up 4-6 weeks after giving birth. Mothers with a lower family income were more likely to not get a postpartum check up. 



## NYC Health Report and Related Work